---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Factor investing

## Factor overview

__Why factor models__:   

- decompose return into explainable and unexplainable components   
- estimate abnormal returns  
- simplify estimate of covariance structure  
- provide portfolio risk analysis for each factor


__Types of factors__:  

- observable (Macroeconomic factor model, Fundamental factor models)
- unobservable (Statistical factor models)


## Macroeconomic factor

$$\underset{\boldsymbol{\alpha},\boldsymbol{\beta}}{\text{minimize}}\quad \sum_{t=1}^T\|\mathbf{x}_t - \boldsymbol{\alpha} - \boldsymbol{\beta}f_t\|^2$$

$$\begin{aligned}
\hat{\boldsymbol{\beta}} & = {\sf cov}(\mathbf{x}_t,f_t)/{\sf var}(f_t)\\
\hat{\boldsymbol{\alpha}} & = \bar{x} - \hat{\boldsymbol{\beta}}\bar{f}\\
\hat{\boldsymbol{\epsilon}}_i & = \mathbf{x}_i - \alpha_i\mathbf{1} - \beta_i\mathbf{f}, \quad i=1,\ldots,N\\
\hat{\sigma}_i^2 & = \frac{1}{T-2}\hat{\boldsymbol{\epsilon}}_i^T\hat{\boldsymbol{\epsilon}}_i, \quad \hat{\boldsymbol{\Psi}} = \sf{diag}(\hat{\sigma}_1^2, \ldots, \hat{\sigma}_N^2)\\
\hat{\boldsymbol{\Sigma}} & = {\sf var}(f_t)\hat{\boldsymbol{\beta}}\hat{\boldsymbol{\beta}}^T + \hat{\boldsymbol{\Psi}}
\end{aligned}$$

__Data preparation__:

```{r message=FALSE, warning=FALSE}
from <- "2017-01-01" # do not change time period for this chapter
to <- "2020-09-30"
SP500_index <- get_data("^GSPC", from, to)

tickers <- c("AAPL", "AMD", "ADI",  "ABBV", "AEZS", "A",  "APD", "AA","CF")
prices <- get_data(tickers, from, to)

X <- diff(log(prices), na.pad = FALSE)
N <- ncol(X)  # number of stocks
T <- nrow(X)  # number of days
f <- diff(log(SP500_index), na.pad = FALSE)
```
### LS fitting -- simple form

```{r}
beta <- cov(X,f)/as.numeric(var(f))
alpha <- colMeans(X) - beta*colMeans(f)
colnames(alpha) <- "alpha"
colnames(beta) <- "beta"
sigma2 <- rep(NA, N)
for (i in 1:N) {
  eps_i <- X[, i] - alpha[i] - beta[i]*f
  sigma2[i] <- (1/(T-2)) * t(eps_i) %*% eps_i
}
Psi <- diag(sigma2)
Sigma <- as.numeric(var(f)) * beta %*% t(beta) + Psi
print(cbind(alpha, beta))
```

### LS fitting -- matrix form

$$
\underset{\boldsymbol{\alpha},\boldsymbol{\beta}}{\text{minimize}}\quad \|\mathbf{X}^T - \boldsymbol{\alpha}\mathbf{1}^T - \boldsymbol{\beta}\mathbf{f}^T\|^2_F. \\
\boldsymbol{\Gamma} = [ \boldsymbol{\alpha}, \boldsymbol{\beta} ] = \mathbf{X}^T \tilde{\mathbf{F}} (\tilde{\mathbf{F}}^T \tilde{\mathbf{F}})^{-1}
$$
where $\tilde{\mathbf{F}} = [\mathbf{1}, \mathbf{f}]$


```{r}
F_ <- cbind(ones = 1, f)
Gamma <- t(X) %*% F_ %*% solve(t(F_) %*% F_)  # better: Gamma <- t(solve(t(F_) %*% F_, t(F_) %*% X))
colnames(Gamma) <- c("alpha", "beta")
alpha <- Gamma[, 1]  # or alpha <- Gamma[, "alpha"]
beta <- Gamma[, 2]   # or beta <- Gamma[, "beta"]
print(Gamma)
E <- xts(t(t(X) - Gamma %*% t(F_)), index(X))  # residuals
Psi <- (1/(T-2)) * t(E) %*% E
Sigma <- as.numeric(var(f)) * beta %o% beta + diag(diag(Psi))
```

### Estimating covriacne

$$
\boldsymbol{\Sigma} = var(f_t)\boldsymbol{\beta}\boldsymbol{\beta}^T + {\sf Diag}(\boldsymbol{\Psi})
$$
where $\boldsymbol{\Psi}$ is residual term

```{r}
library(corrplot)  
corrplot(cov2cor(Sigma),  
         type="upper", 
         tl.col="black", tl.srt=45,     #Text label color and rotation
         diag=FALSE)
```

```{r}
corrplot(cov2cor(Psi), mar = c(0,0,1,0), order = "hclust", addrect = 3, 
        title = "Covariance matrix of residuals", diag = FALSE, tl.col="black", tl.srt=45)
```

Interestingly, we can observe that the automatic clustering performed on $\boldsymbol{\Psi}$ correctly identifies the sectors of the stocks (information tech, health care and materials).

### Evaluating ETFs performance
In this section, we will use the S&P 500 as the explicit market factor and assume risk free rate equal to 0. 

Consider following six ETFs:

- SPY - SPDR S&P 500 ETF (index tracking)  
- SPHB - PowerShares S&P 500 High Beta Portfolio (high beta)  
- SPLV - PowerShares S&P 500 Low Volatility Portfolio (low beta)  
- USMV - iShares Edge MSCI Min Vol USA ETF  
- JKD - iShares Morningstar Large-Cap ETF

```{r message=FALSE, warning=FALSE}
index <- c("SPY","SPHB", "SPLV", "USMV", "JKD")
ETFs <- get_data(index, from, to)
X_ETFs <- diff(log(ETFs), na.pad = FALSE)

# compute the alpha and beta of all the ETFs:
F_ <- cbind(ones = 1, f)
Gamma <- t(solve(t(F_) %*% F_, t(F_) %*% X_ETFs))
colnames(Gamma) <- c("alpha", "beta")
alpha <- Gamma[, 1]
beta <- Gamma[, 2]
print(Gamma)
# visualization of Gamma
{ par(mfrow = c(1,2))    # two plots side by side
  barplot(rev(alpha), horiz = TRUE, main = "alpha", border = NA, col = "coral", cex.names = 0.75, las = 1)
  barplot(rev(beta), horiz = TRUE, main = "beta", border = NA, col = "deepskyblue", cex.names = 0.75, las = 1)}  # reset to normal single plot
```
`TODO Explanation for the above Gamma graph for all 6 etfs`

Sharpe ratio follows:

$$SR = \frac{E[x_t]}{\sqrt{{\sf var}[x_t]}} = \frac{\alpha + \beta E[f_t]}{\sqrt{\beta^2 {\sf var}[f_t] + \sigma^2}} = \frac{\alpha/\beta + E[f_t]}{\sqrt{{\sf var}[f_t] + \sigma^2/\beta^2}} \approx \frac{\alpha/\beta + E[f_t]}{\sqrt{{\sf var}[f_t]}}$$
```{r}
idx_sorted <- sort(alpha/beta, decreasing = TRUE, index.return = TRUE)$ix
SR <- colMeans(X)/sqrt(diag(var(X)))
ranking <- cbind("alpha/beta" = (alpha/beta)[idx_sorted], 
                 SR = SR[idx_sorted], 
                 alpha = alpha[idx_sorted], 
                 beta = beta[idx_sorted])
print(ranking)
```
`TODO Explain the resul`

## Fama-French
Data preparation:

```{r}
# download Fama-French factors from website
# url <- "http://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_Factors_daily_CSV.zip"
# temp <- tempfile()
# download.file(url, temp, method = "libcurl", mode = "wb")
# unzip(temp, "F-F_Research_Data_Factors_daily.CSV")
# unlink(temp)

mydata <- read.csv("F-F_Research_Data_Factors_daily.CSV", skip = 4)
mydata <- mydata[-nrow(mydata), ]  # remove last row
fama_lib <- xts(x = mydata[, c(2,3,4)], order.by = as.Date(paste(mydata[, 1]), "%Y%m%d"))

# compute the log-returns of the stocks and the Fama-French factors
F <- fama_lib[index(X)]/100
```

$$\underset{\boldsymbol{\alpha},\boldsymbol{\beta}}{\text{minimize}}\quad \|\mathbf{X}^T - \boldsymbol{\Gamma}\tilde{\mathbf{F}}^T\|^2_F$$
$$\boldsymbol{\Gamma} = \mathbf{X}^T \tilde{\mathbf{F}} (\tilde{\mathbf{F}}^T \tilde{\mathbf{F}})^{-1}$$
```{r}
F_ <- cbind(ones = 1, F)
Gamma <- t(solve(t(F_) %*% F_, t(F_) %*% X))
colnames(Gamma) <- c("alpha", "b1", "b2", "b3")
alpha <- Gamma[, 1]
B <- Gamma[, 2:4]
print(Gamma)
```

## Statistical/principal factor models
Recall the principal factor method for the model $\mathbf{X}^T = \boldsymbol{\alpha}\mathbf{1}^T + \mathbf{B}\mathbf{F}^T + \mathbf{E}^T$ with K factors:

1. PCA:

- sample mean: $\hat{\boldsymbol{\alpha}} = \bar{\mathbf{x}} = \frac{1}{T}\mathbf{X}^T\mathbf{1}_T$
- demeaned matrix: $\bar{\mathbf{X}} = \mathbf{X} - \mathbf{1}_T\bar{\mathbf{x}}^T$
- eigen-decomposition: $\hat{\boldsymbol{\Sigma}} = \hat{\boldsymbol{\Gamma}} \hat{\boldsymbol{\Lambda}} \hat{\boldsymbol{\Gamma}}^T$

2. Estimates:

- $\hat{\mathbf{B}} = \hat{\boldsymbol{\Gamma}_1} \hat{\boldsymbol{\Lambda}}_1^{1/2}$
- $\hat{\boldsymbol{\Psi}} = {\sf Diag}\left(\hat{\boldsymbol{\Sigma}} - \hat{\mathbf{B}} \hat{\mathbf{B}}^T\right)$
- $\hat{\boldsymbol{\Sigma}} = \hat{\mathbf{B}} \hat{\mathbf{B}}^T + \hat{\boldsymbol{\Psi}}$

3. Update the eigen-decomposition: $\hat{\boldsymbol{\Sigma}} - \hat{\boldsymbol{\Psi}} = \hat{\boldsymbol{\Gamma}} \hat{\boldsymbol{\Lambda}} \hat{\boldsymbol{\Gamma}}^T$

4. Repeat Steps 2-3 until convergence.

```{r}
K <- 3
alpha <- colMeans(X)
X_ <- X - matrix(alpha, T, N, byrow = TRUE)
Sigma_prev <- matrix(0, N, N)
Sigma <- (1/(T-1)) * t(X_) %*% X_
eigSigma <- eigen(Sigma)
while (norm(Sigma - Sigma_prev, "F")/norm(Sigma, "F") > 1e-3) {
  B <- eigSigma$vectors[, 1:K, drop = FALSE] %*% diag(sqrt(eigSigma$values[1:K]), K, K)
  Psi <- diag(diag(Sigma - B %*% t(B)))
  Sigma_prev <- Sigma
  Sigma <- B %*% t(B) + Psi
  eigSigma <- eigen(Sigma - Psi)
}
cbind(alpha, B)
```
## Comparison of covariance matrix estimations

```{r}
# prepare Fama-French factors
F_FamaFrench <- F # rename for easily knowing the meaning of variable
# prepare index
f_SP500 <- f # rename for easily knowing the meaning of variable

# split data into training and set data
T_trn <- round(0.45*T)
X_trn <- X[1:T_trn, ]
X_tst <- X[(T_trn+1):T, ]
F_FamaFrench_trn <- F_FamaFrench[1:T_trn, ]
F_FamaFrench_tst <- F_FamaFrench[(T_trn+1):T, ]
f_SP500_trn <- f_SP500[1:T_trn, ]
f_SP500_tst <- f_SP500[(T_trn+1):T, ]
```


```{r}
# sample covariance matrix
Sigma_SCM <- cov(X_trn)

# 1-factor model
F_ <- cbind(ones = 1, f_SP500_trn)
Gamma <- t(solve(t(F_) %*% F_, t(F_) %*% X_trn))
colnames(Gamma) <- c("alpha", "beta")
alpha <- Gamma[, 1]
beta <- Gamma[, 2]
E <- xts(t(t(X_trn) - Gamma %*% t(F_)), index(X_trn))
Psi <- (1/(T_trn-2)) * t(E) %*% E
Sigma_SP500 <- as.numeric(var(f_SP500_trn)) * beta %o% beta + diag(diag(Psi))

# Fama-French 3-factor model
F_ <- cbind(ones = 1, F_FamaFrench_trn)
Gamma <- t(solve(t(F_) %*% F_, t(F_) %*% X_trn))
colnames(Gamma) <- c("alpha", "beta1", "beta2", "beta3")
alpha <- Gamma[, 1]
B <- Gamma[, 2:4]
E <- xts(t(t(X_trn) - Gamma %*% t(F_)), index(X_trn))
Psi <- (1/(T_trn-4)) * t(E) %*% E
Sigma_FamaFrench <- B %*% cov(F_FamaFrench_trn) %*% t(B) + diag(diag(Psi))

# Statistical 1-factor model
K <- 1
alpha <- colMeans(X_trn)
X_trn_ <- X_trn - matrix(alpha, T_trn, N, byrow = TRUE)
Sigma_prev <- matrix(0, N, N)
Sigma <- (1/(T_trn-1)) * t(X_trn_) %*% X_trn_
eigSigma <- eigen(Sigma)
while (norm(Sigma - Sigma_prev, "F")/norm(Sigma, "F") > 1e-3) {
  B <- eigSigma$vectors[, 1:K, drop = FALSE] %*% diag(sqrt(eigSigma$values[1:K]), K, K)
  Psi <- diag(diag(Sigma - B %*% t(B)))
  Sigma_prev <- Sigma
  Sigma <- B %*% t(B) + Psi
  eigSigma <- eigen(Sigma - Psi)
}
Sigma_PCA1 <- Sigma

# Statistical 3-factor model
K <- 3
alpha <- colMeans(X_trn)
X_trn_ <- X_trn - matrix(alpha, T_trn, N, byrow = TRUE)
Sigma_prev <- matrix(0, N, N)
Sigma <- (1/(T_trn-1)) * t(X_trn_) %*% X_trn_
eigSigma <- eigen(Sigma)
while (norm(Sigma - Sigma_prev, "F")/norm(Sigma, "F") > 1e-3) {
  B <- eigSigma$vectors[, 1:K] %*% diag(sqrt(eigSigma$values[1:K]), K, K)
  Psi <- diag(diag(Sigma - B %*% t(B)))
  Sigma_prev <- Sigma
  Sigma <- B %*% t(B) + Psi
  eigSigma <- eigen(Sigma - Psi)
}
Sigma_PCA3 <- Sigma

# Statistical 5-factor model
K <- 5
alpha <- colMeans(X_trn)
X_trn_ <- X_trn - matrix(alpha, T_trn, N, byrow = TRUE)
Sigma_prev <- matrix(0, N, N)
Sigma <- (1/(T_trn-1)) * t(X_trn_) %*% X_trn_
eigSigma <- eigen(Sigma)
while (norm(Sigma - Sigma_prev, "F")/norm(Sigma, "F") > 1e-3) {
  B <- eigSigma$vectors[, 1:K] %*% diag(sqrt(eigSigma$values[1:K]), K, K)
  Psi <- diag(diag(Sigma - B %*% t(B)))
  Sigma_prev <- Sigma
  Sigma <- B %*% t(B) + Psi
  eigSigma <- eigen(Sigma - Psi)
}
Sigma_PCA5 <- Sigma
```

```{r}
Sigma_true <- cov(X_tst)
error <- c(SCM = norm(Sigma_SCM - Sigma_true, "F"),
           SP500 = norm(Sigma_SP500 - Sigma_true, "F"),
           FamaFrench = norm(Sigma_FamaFrench - Sigma_true, "F"),
           PCA1 = norm(Sigma_PCA1 - Sigma_true, "F"),
           PCA3 = norm(Sigma_PCA3 - Sigma_true, "F"),
           PCA5 = norm(Sigma_PCA5 - Sigma_true, "F"))
print(error)
barplot(error, main = "Error in estimation of covariance matrix (Frobenius norm)", 
        col = "aquamarine3", cex.names = 0.75, las = 1)
```

$${\sf PRIAL} = 100\times \frac{\|\boldsymbol{\Sigma}_{\sf scm} - \boldsymbol{\Sigma}_{\sf true}\|_F^2 - \|\hat{\boldsymbol{\Sigma}} - \boldsymbol{\Sigma}_{\sf true}\|_F^2}{\|\boldsymbol{\Sigma}_{\sf scm} - \boldsymbol{\Sigma}_{\sf true}\|_F^2}$$
```{r}
ref <- norm(Sigma_SCM - Sigma_true, "F")^2
PRIAL <- 100*(ref - error^2)/ref
print(PRIAL)
barplot(PRIAL, main = "PRIAL for estimation of covariance matrix", 
        col = "bisque3", cex.names = 0.75, las = 1)
```
The final conclusion is that using factor models for covariance matrix estimation may help. In addition, it is not clear that using explicit factors helps as compared to the statistical factor modeling.


## Edhec smart beta