# Alternative risk measure

## Motivation {#motivation_alter_risk_measure}

One of main objective of this book is improving the 3 limits (risk measure, sensitivity to estimation errors and lacks of risk diversification) of Markowitz's model `TODO REF FORMER CHARPTER`.

Variance is not a good measure of risk in practice since it penalizes both the unwanted high losses and the desired low losses (or gains)




We will here consider more meaningful measures for risk than the variance, like the downside risk (DR), Value-at-Risk (VaR), Conditional VaR (CVaR) or Expected Shortfall (ES), and drawdown (DD).

## Risk measure

The mean return is not good enough and one needs to control the probability of going bankrupt. To overcome the limitations of the variance as risk measure, a number of alternative risk measures have been proposed, for example:  

- Downside Risk (DR)  
- Value-at-Risk (VaR)  
- Conditional Value-at-Risk (CVaR)  
- Drawdown (DD):  
    - maximum DD  
    - average DD  
    - Conditional Drawdown at Risk (CDaR)  


```{r message=FALSE, warning=FALSE}
from = "2013-01-01"
to = "2016-12-31"
tickers <- c("AAPL", "AMD", "ADI",  "ABBV", "AEZS", "A",  "APD", "AA","CF")
assign_variable_global_env(tickers, from, to, ratio = 0.7)

N <- ncol(prices)
T <- nrow(prices)
T_trn <- round(T*0.7)
X <- X_log

mu <- colMeans(X)
Sigma <- cov(X)
plot(prices/rep(prices[1, ], each = nrow(prices)), col = rainbow10equal, legend.loc = "topleft",main = "Normalized prices")
```

```{r}
w_Markowitz <- MVP(mu, Sigma, lmd = 0.5, long_only = TRUE)
w_GMVP <- GMVP(Sigma, long_only = TRUE)
w_all <- cbind("GMVP"      = w_GMVP, 
               "Markowitz" = w_Markowitz)
barplot(t(w_all), col = rainbow10equal[1:2], legend = colnames(w_all), beside = TRUE,
        main = "Portfolio allocation", xlab = "stocks", ylab = "dollars")

# compute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_all_trn))
t(table.AnnualizedReturns(ret_all_tst))
{ chart.CumReturns(ret_all, main = "Cumulative return of portfolios", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich10equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
chart.CumReturns(ret_all_tst, main = "Cumulative return of portfolios (out-of-sample)",
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich10equal)
{ chart.Drawdown(ret_all, main = "Drawdown of portfolios", 
                 legend.loc = "bottomleft", colorset = rich8equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
```

## Alternative risk measure beased portfolio

### Mean-downside risk portfolio

$$\begin{array}{ll}
\underset{\mathbf{w}}{\textsf{maximize}} & \mathbf{w}^T\boldsymbol{\mu}-\lambda \frac{1}{T}\sum_{t=1}^T\left(\left(\tau - \mathbf{w}^T\mathbf{r}_t\right)^+\right)^\alpha\\
\textsf{subject to} & \mathbf{1}^T\mathbf{w}=1,\quad \mathbf{w}\ge\mathbf{0}.
\end{array}$$

The semi-variance is a special case of the more general lower partial moments (LPM)
$$\textsf{LPM} = \mathsf{E}\left[\left((\tau - R)^+\right)^\alpha\right]$$
where $(\cdot)^+=\max(0, \cdot)$

The parameter α reflects the investor’s feeling about the relative consequences of falling short of τ by various amounts: the value α = 1 (which suits a neutral investor) separates risk-seeking ( 0 < α < 1 ) from risk-averse ( α > 1 ) behavior with regard to returns below the target τ.

```{r}
w_DR_alpha1 <- portfolioDR(X_log_trn, alpha = 1)
w_DR_alpha2 <- portfolioDR(X_log_trn, alpha = 2)
w_DR_alpha3 <- portfolioDR(X_log_trn, alpha = 3)

# combine portfolios
w_all <- cbind(
  w_all,
  "DR-alpha-1" = w_DR_alpha1,
  "DR-alpha-2" = w_DR_alpha2,
  "DR-alpha-3" = w_DR_alpha3
)

# compute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_all_trn))
t(table.AnnualizedReturns(ret_all_tst))
table.DownsideRisk(ret_all_tst)
{ chart.CumReturns(ret_all, main = "Cumulative return of portfolios", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich6equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
chart.CumReturns(ret_all_tst, main = "Cumulative return of portfolios (out-of-sample)", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich6equal)

{ chart.Drawdown(ret_all, main = "Drawdown of portfolios", 
                 legend.loc = "bottomleft", colorset = rich6equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
w_all <- w_all[, ! colnames(w_all) %in% c("DR-alpha-1", "DR-alpha-2")]
```

### Mean-CVaR portfolio
$$\begin{array}{ll}
\underset{\mathbf{w}}{\textsf{maximize}} & \mathbf{w}^{T}\boldsymbol{\mu}\\
\textsf{subject to} 
 & \mathsf{CVaR}_{\alpha}\left(f\left(\mathbf{w},\mathbf{r}\right)\right)\leq c\\
 & \mathbf{1}^T\mathbf{w}=1,\quad \mathbf{w}\ge\mathbf{0}
\end{array}$$

$$\begin{array}{ll}
\underset{\mathbf{w}, \mathbf{z}, \zeta}{\textsf{maximize}} & \mathbf{w}^T\boldsymbol{\mu} - \lambda\left(\zeta+\frac{1}{1-\alpha}\frac{1}{T}\sum_{t=1}^{T}z_{t}\right)\\
\textsf{subject to} 
 & 0\leq z_{t}\geq-\mathbf{w}^{T}\mathbf{r}_{t}-\zeta,\quad t=1,\dots,T\\
 & \mathbf{1}^T\mathbf{w}=1,\quad \mathbf{w}\ge\mathbf{0}.
\end{array}$$

```{r}
w_CVaR095 <- portolioCVaR(X_log_trn, alpha = 0.95)
w_CVaR099 <- portolioCVaR(X_log_trn, alpha = 0.99)

# combine portfolios
w_all <- cbind(w_all, 
               "CVaR-alpha-0.95" = w_CVaR095,
               "CVaR-alpha-0.99" = w_CVaR099)

# compute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_all_trn))
t(table.AnnualizedReturns(ret_all_tst))
{ chart.CumReturns(ret_all, main = "Cumulative return of portfolios", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich6equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
chart.CumReturns(ret_all_tst, main = "Cumulative return of portfolios (out-of-sample)", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich6equal)
{ chart.Drawdown(ret_all, main = "Drawdown of portfolios", 
                 legend.loc = "bottomleft", colorset = rich6equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
w_all <- w_all[, ! colnames(w_all) %in% c("CVaR-alpha-0.95")]
```


### Mean - Max-DD portfolio

$$D(t)=\max_{1\le\tau\le t}r_p^{\sf cum}(\tau) - r_p^{\sf cum}(t)$$

- maximum DD (Max-DD) 
- average DD (Ave-DD) 
- Conditional Drawdown at Risk (CDaR)

$$\begin{array}{ll}
\underset{\mathbf{w}, \{u_t\}}{\textsf{maximize}} & \mathbf{w}^T\boldsymbol{\mu}\\
\textsf{subject to} 
 & \mathbf{w}^T\mathbf{r}_{t}^{\sf cum} \le u_t \le \mathbf{w}^T\mathbf{r}_t^{\sf cum} + c, \quad\forall 1\le t\le T\\
 & u_{t-1} \le u_t\\
 & \mathbf{1}^T\mathbf{w}=1,\quad \mathbf{w}\ge\mathbf{0}.
\end{array}$$


```{r}
# solver may not find a solution for some c if c is too low
w_MaxDD_c018 <- portfolioMaxDD(X_log_trn, c = 0.18)
w_MaxDD_c021 <- portfolioMaxDD(X_log_trn, c = 0.21)
w_MaxDD_c024 <- portfolioMaxDD(X_log_trn, c = 0.24)

# combine portfolios
w_all <- cbind(w_all,
               "Max-DD-c-018" = w_MaxDD_c018,
               "Max-DD-c-021" = w_MaxDD_c021,
               "Max-DD-c-024" = w_MaxDD_c024)

# compute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_all_tst)[3, ])
t(maxDrawdown(ret_all_trn))
t(maxDrawdown(ret_all_tst))
{ chart.CumReturns(ret_all, main = "Cumulative return of portfolios",
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich8equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
chart.CumReturns(ret_all_tst, main = "Cumulative return of portfolios (out-of-sample)",
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich8equal)
{ chart.Drawdown(ret_all, main = "Drawdown of portfolios",
                 legend.loc = "bottomleft", colorset = rich8equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
w_all <- w_all[, ! colnames(w_all) %in% c("Max-DD-c-018", "Max-DD-c-024")]
```

### Mean - Ave-DD portfolio

$$\begin{array}{ll}
\underset{\mathbf{w}, \{u_t\}}{\textsf{maximize}} & \mathbf{w}^T\boldsymbol{\mu}\\
\textsf{subject to} 
 & \frac{1}{T}\sum_{t=1}^T u_t \le \sum_{t=1}^T\mathbf{w}^T\mathbf{r}_t^{\sf cum} + c\\
 & \mathbf{w}^T\mathbf{r}_{t}^{\sf cum} \le u_t\\
 & u_{t-1} \le u_t\\
 & \mathbf{1}^T\mathbf{w}=1,\quad \mathbf{w}\ge\mathbf{0}.
\end{array}$$

```{r}
w_AveDD_c004 <- portfolioAveDD(X_log_trn, c = 0.04)
w_AveDD_c006 <- portfolioAveDD(X_log_trn, c = 0.06)
w_AveDD_c008 <- portfolioAveDD(X_log_trn, c = 0.08)

# combine portfolios
w_all <- cbind(w_all, 
               "Ave-DD-c-004" = w_AveDD_c004, 
               "Ave-DD-c-006" = w_AveDD_c006, 
               "Ave-DD-c-008" = w_AveDD_c008)

# compute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_all_tst)[3, ])
t(maxDrawdown(ret_all_tst))
t(AverageDrawdown(ret_all_trn))
t(AverageDrawdown(ret_all_tst))
{ chart.CumReturns(ret_all, main = "Cumulative return of portfolios", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich8equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
chart.CumReturns(ret_all_tst, main = "Cumulative return of portfolios (out-of-sample)", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich8equal)
{ chart.Drawdown(ret_all, main = "Drawdown of portfolios", 
                 legend.loc = "bottomleft", colorset = rich8equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
w_all <- w_all[, ! colnames(w_all) %in% c("Ave-DD-c-006", "Ave-DD-c-008")]
```

### Mean-CDaR portfolio
$$\begin{array}{ll}
\underset{\mathbf{w}, \{z_t\}, \zeta, \{u_t\}}{\textsf{maximize}} & \mathbf{w}^T\boldsymbol{\mu}\\
\textsf{subject to} 
 & \zeta+\frac{1}{1-\alpha}\frac{1}{T}\sum_{t=1}^{T}z_{t} \le c\\
 & 0\leq z_{t}\geq u_t - \mathbf{w}^T\mathbf{r}_t^{\sf cum} - \zeta, \quad t=1,\dots,T\\
 & \mathbf{w}^T\mathbf{r}_{t}^{\sf cum} \le u_t\\
 & u_{t-1} \le u_t\\
 & \mathbf{1}^T\mathbf{w}=1,\quad \mathbf{w}\ge\mathbf{0}.
\end{array}$$

```{r}
w_CDaR095_c014 <- portfolioCDaR(X_log_trn, c = 0.14, alpha = 0.95)
w_CDaR095_c016 <- portfolioCDaR(X_log_trn, c = 0.16, alpha = 0.95)
w_CDaR099_c017 <- portfolioCDaR(X_log_trn, c = 0.17, alpha = 0.99)
w_CDaR099_c019 <- portfolioCDaR(X_log_trn, c = 0.19, alpha = 0.99)

# combine portfolios
w_all <- cbind(w_all, 
               "CDaR095-c-014" = w_CDaR095_c014, 
               "CDaR095-c-016" = w_CDaR095_c016, 
               "CDaR099-c-017" = w_CDaR099_c017, 
               "CDaR099-c-019" = w_CDaR099_c019)

# compute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_all_tst)[3, ])
t(maxDrawdown(ret_all_tst))
t(AverageDrawdown(ret_all_tst))
t(CDD(ret_all_trn))
t(CDD(ret_all_tst))
{ chart.CumReturns(ret_all, main = "Cumulative return of portfolios", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich10equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
chart.CumReturns(ret_all_tst, main = "Cumulative return of portfolios (out-of-sample)", 
                   wealth.index = TRUE, legend.loc = "topleft", colorset = rich10equal)
{ chart.Drawdown(ret_all, main = "Drawdown of portfolios", 
                 legend.loc = "bottomleft", colorset = rich10equal)
  addEventLines(xts("training", index(X_lin[T_trn])), srt=90, pos=2, lwd = 2, col = "darkblue") }
w_all <- w_all[, ! colnames(w_all) %in% c("CDaR095-c-016", "CDaR099-c-017", "CDaR099-c-019")]
```

## Comparison of DR, CVaR, and DD portfolios

```{r out.width='100%'}
# recompute returns of all portfolios
ret_all <- xts(X_lin %*% w_all, index(X_lin))
ret_all_trn <- ret_all[1:T_trn, ]
ret_all_tst <- ret_all[-c(1:T_trn), ]

t(table.AnnualizedReturns(ret_all_tst)[3, ])
t(maxDrawdown(ret_all_tst))
t(AverageDrawdown(ret_all_tst))
t(CDD(ret_all_tst))
{par(mfrow = c(1, 1), mar = c(5, 5, 4, 3))
barplot(w_all, col = rainbow10equal[1:9], legend = rownames(w_all), args.legend = list(x="topright", bty = "n", cex=0.8, horiz=FALSE, inset = c(-0.1,0)), beside = TRUE, main = "Portfolio allocation", xlab = "portfolio design", ylab = "weights", las=2, cex.names=.6)}
```

## Conclusion {#alter_con}

The maximum drawdown is extremely sensitive to minute changes in the portfolio weights and to the specific time period examined.

If the returns are close to normally distributed, the distribution of drawdowns is just a function of the variance, so there’s no need to include drawdowns explicitly in your portfolio construction objective. Minimizing variance is the same as minimizing expected drawdowns.

On the other hand, if returns are very non-normal and you want to find a portfolio that minimizes the expected drawdowns, you still wouldn’t choose weights that minimize historical drawdown. Why?

Because minimizing historical drawdown is effectively the same as taking all your returns that weren’t part of a drawdown, and hiding them from your optimizer, which will lead to portfolio weights that are a lot less accurately estimated than if you let your optimizer see all the data you have.

Instead, you might just include terms in your optimization objective that penalize negative skew and penalize positive kurtosis 

As a note of caution, one has to be careful with CVaR and CDaR portfolios due to the sensitivity to the used data (if not enough data, they will not be reliable).


